Best Practices in Latent Class Analysis using Free Open Source Software

### 2023-01-23 manuscript.Rmd
edits by Mauricio

---
title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "c.j.vanlissa@gmail.com"
  - name          : "Mauricio Garnier-Villarreal"
    affiliation   : "3"
    corresponding : no   
    email         : "m.garniervillarreal@vu.nl"
  - name          : "Daniel Anadria"
    affiliation   : "1"
    corresponding : no
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "d.anadria@uu.nl"
affiliation:
  - id            : "1"
    institution   : "Utrecht University, Methodology & Statistics"
  - id            : "2"
    institution   : "Open Science Community Utrecht"
  - id            : "3"
    institution   : "Vrije Universiteit Amsterdam, Sociology"
author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands
abstract: |
  Latent class analysis (LCA) refers to a family of techniques for identifying groups
  in data based on a parametric model.
  Examples include mixture models, LCA with ordinal
  indicators, and latent class growth analysis.
  Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of LCA,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package whose backend relies on `OpenMx`.
keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
wordcount         : "9001"
bibliography      : "mixture.bib"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
class             : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: 72
---

@@

Despite the popularity of LCA,
two challenges impede broader adoption and correct application of the method.
The first challenge is that most existing software that implements these models is commercial and closed-source.
The second challenge is that there is a lack of standards for estimating and reporting LCA, and some myths and misunderstandings have been perpetuated in prior literature.
This introduces a risk of misapplications of the technique,
and complicates manuscript review and assessment of the quality of published studies.
The present paper seeks to address these challenges by introducing updated guidelines for estimatin and reporting LCA,
based on current best practice. 
Furthermore, meeting the increasing demand for open source research software,
this paper introduces new functionality in the R-package `tidySEM` [@vanlissaTidySEMTidyStructural2022] to perform LCA using free open source software (FOSS).

@@

There are several compelling reasons to prefer free open source software (FOSS) instead.
FOSS is freely available for anyone to use, modify, and distribute, which means that scientists can access and use a wide range of powerful tools without financial impediments.
This is invaluable for reducing gatekeeping of researchers with limited budgets, such as those from developing countries.
Furthermore, FOSS promotes collaboration within the scientific community,
as its users can contribute to the development, maintenance, and support of the software.
Finally, FOSS promotes tranparency and reproducibility: Because the source code for FOSS is accessible,
researchers can review and verify the methods and algorithms used,
which helps to ensure the integrity and reliability of research findings.
Overall, FOSS has the potential to significantly improve the efficiency and effectiveness of scientific research, while also promoting transparency, collaboration, and inclusivity within the scientific community.

Notable FOSS for LCA includes Mclust [REF], depmixS4 [REF], and OpenMx [@neale_openmx_2016].
A crucial limitation of these existing packages is that some have limited functionality (e.g., Mclust),
or implement specific innovations (e.g., depmixS4).
OpenMx is unique in that, like Mplus, it implement LCA in the context of a fully featured structural equation modeling framework.
Remaining obstacles to the use of `OpenMx` is that it is not very user-friendly and its functionality for LCA is poorly documented.

@@

A common type of LCA is the *finite Gaussian mixture model*; a
univariate analysis where the observed distribution of a single variable
is assumed to result from a mixture of a known number of normal
(Gaussian) distributions. The parameters of a finite Gaussian mixture
model are the means and variances of these underlying normal
distributions. The analysis of shoe sizes presented earlier is a
canonical example of this type of analysis. In the multivariate case,
with more than one indicator variable, the parameters of a mixture model
are the means, variances, and covariances between the indicators (which
can be standardized to obtain correlations). These parameters can be estimated 
freely, or set to be constrained across classes [@collins_latent_2009].

@@

@@

Finally, LCA can be used to deal with data which violate certain assumptions.
For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughmanMixtureModelFramework2007].
These models estimate one class to explain the "true zeros",
and another class with a different distribution to explain all oter values, which can include zero.
Note that this LCA has different models for both classes,
and thus falls beyond the scope of this paper.
<!-- CJ: This has been said elsewhere: -->
<!-- In fact, LCA assumes the population -->
<!-- distribution is a non-normal mixture of $K$ normal distributions, and it -->
<!-- can recover the value of $K$, i.e. generate a $K$-class solution. -->
<!-- CJ: This example is not perfect, we need another example still.. -->
For instance, the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality [@alkema_maximizing_2022].
A mixture of two normal distributions with distinct variances described the observed distribution better.
<!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical. We encourage researchers to be agnostic about the **true** nature of the latent variable. And focus on the approach that fits the research question. -->
<!-- CJ: Why would we encourage researchers to be agnostic about the true...? -->

<!-- ### Controlling for covariates -->
<!-- CJ: I'm not sure we should call this a use case... in principle, -->
<!-- any time you're doing some kind of LCA you may want to predict group membership or predict something from group membership... -->
<!-- An extension of LCA is that containing covariates which can be used to -->
<!-- predict class membership. In this approach, we not only model the latent -->
<!-- class variable based on indicators, but we also relate the class -->
<!-- membership to other explanatory variables (@vermunt_latent_2017). An -->
<!-- example of using covariates comes from @nozadi_moderating_2016 who -->
<!-- applied LCA to identify the probability of children's membership to an -->
<!-- anxiety class. The authors tested several covariates including -->
<!-- children's age, sex, and accuracy scores. Age and sex were not found to -->
<!-- be related to the children's latent class membership, hence these -->
<!-- covariates were excluded from the analysis. Accuracy scores, however, -->
<!-- were related to probabilities of being in anxiety and attention-anxiety -->
<!-- classes. Therefore this covariate was kept as a valuable predictor. -->
<!-- Another example comes from @van_lissa_divergence_2015 who included sex  -->
<!-- as a predictor of the developmental trajectories and class membership in  -->
<!-- the previously discussed adolescent empathy development study. The rationale -->
<!-- was that adding sex as a covariate would account for known sex differences  -->
<!-- in empathy development, and lead to a model with lower bias. -->

<!-- ### Distal outcomes -->
<!-- CJ: I'm not sure we should call this a use case... in principle, -->
<!-- any time you're doing some kind of LCA you may want to predict group membership or predict something from group membership... -->
<!-- When our interest is the prediction of one or more outcomes, LCA can be used  -->
<!-- to construct latent classes as categorical predictors. @lanza_latent_2013  -->
<!-- demonstrated how LCA can be used to classify adolescents into depression  -->
<!-- classes, and subsequently these classes can be used to predict smoking, grades, -->
<!-- and delinquency. The study showed that the outcomes predicted by class  -->
<!-- membership can be binary (regular smoking), continuous (grades) or  -->
<!-- count (delinquency). In the previously discussed study of Belgian adolescent -->
<!-- identities, @maene_perceived_2022 first constructed the latent variable -->
<!-- representing different multiple identification strategies. Subsequently,  -->
<!-- the most likely class membership based on posterior class probability was -->
<!-- used as an "observed" categorical variable for prediction of depressive symptoms. -->

@@

<!-- Any variables that of class `character` or `factor` must be recoded to ordinal. -->
<!-- Nominal variables must be recoded to multiple ordinal dummy variables, e.g., using the function `mx_dummies()`.  -->
LCA can accommodate variables of different measurement levels: continuous, binary, nominal, and ordinal.
Numeric variables should have an interval measurement level,
such that it is sensible to estimate means and (co)variances.
For the remaining three measurement levels,
some preprocessing is required.
The exact data types accepted by `tidySEM` at the time of writing are `numeric` and `ordered`.
To correctly specify ordered variables
such as Likert scales,
use the function `OpenMx::mxFactor()`.
Binary variables,
such as presence or absence of some symptom, are also treated as ordinal, with one category coded as zero and the other as one.
Nominal variables are converted to binary indicators of group membership via dummy coding, for example, using `mx_dummies()`.

@@

Previous work has emphasized the importance of examining the pattern of
missing data and reporting how missingness was handled
[@van_de_schoot_grolts-checklist_2017].
While we concur that investigating missingness is due diligence,
it is important to emphasize that missingness is adequately handled by default in `OpenMx`, which is the computational backend of `tidySEM`.
Its Full Information Maximum Likelihood (FIML) estimator makes use of all available
information without imputing missing values [@endersRelativePerformanceFull2001].
FIML is a best-practice solution for handling missing data;
at least on par with multiple imputation [@lee_comparison_2021].
Multiple imputation is less suitable to LCA for two reasons. First,
because LCA methods are often computationally intensive,
and conducting them on several imputed datasets multiplies the load.
Second, because many multiple imputation methods assume normality; an assumption likely to be violated when data come from a heterogeneous population.

@@

<!-- In the general sense, the class-specific model can be any structural equation model. -->
<!-- Its parameters may include means, (co)variances, regression coefficients, factor loadings, and thresholds for ordinal variables. -->
<!-- All of these parameters can be freely estimated, -->
<!-- constrained to be equal across classes, -->
<!-- or fixed to a known value. --> 
<!-- The specific types LCA introduced in the Taxonomy all have different parameters. -->
<!-- For example, LPA typically assumes class-specific means, -->
<!-- equal variances across classes, and zero covariances (i.e., indicators are independent after accounting for latent class membership), see `mx_profiles()`. -->
<!-- By contrast, LCA with ordinal indicators assumes free thresholds across classes, -->
<!-- where each threshold reflects a proportion of responses within a particular response category, see `mx_lca()`. -->
<!-- Like LPA, LCGA and GMM typically have free means across classes, -->
<!-- but with respect to the latent growth variables instead of the observed variables, see `mx_growth_mixture()`. -->
<!-- The distinction between LCGA and GMM is whether variances of the latent growth variables are constrained to zero, or freely estimated (either across or within classes). -->
<!-- either using `lavaan` model syntax or using `OpenMx` RAM specification, -->
<!-- and supplied to the general function `mx_mixture()`. -->
<!-- The resulting model may have -->
<!-- HIER -->
<!-- For latent profile analysis, the function -->
<!-- `tidySEM::mx_profiles(classes, variances, covariances)` largely -->
<!-- automates this process. The argument `classes` indicates which class -->
<!-- solutions should be estimated (e.g., 1 through 6). The argument -->
<!-- `variances` specifies whether variances should be `"equal"` or -->
<!-- `"varying"` across classes. The argument `covariances` specifies whether -->
<!-- covariances should be constrained to `"zero"`, `"equal"` or `"varying"` -->
<!-- across classes. The means are free to vary across classes by default, -->
<!-- although the more general function `tidySEM::mx_mixture()` could be used -->
<!-- to circumvent this. After all models have been estimated, the function -->
<!-- `tidySEM::table_fit()` can be used to obtain a model fit table suitable -->
<!-- for determining the optimal model according to best practices. Note, -->
<!-- however, that this table does not include the bootstrapped likelihood -->
<!-- ratio test (BLRT) by default, because this test is very computationally -->
<!-- expensive. It is recommended to use the function `tidySEM::BLRT()` to -->
<!-- compare a shortlist of likely candidate models based on other fit -->
<!-- indices. More on fit indices can be found in the Model Fit Indices -->
<!-- subsection of this paper. -->

@@

<!-- A problem which can occur is that of inadmissible solutions. With binary -->
<!-- indicators, LCA is modeling a cross-table with all the predictors. The -->
<!-- problem with such cross-tables is that they will often contain empty -->
<!-- cells, i.e. combinations of responses that never occur together. This -->
<!-- problem is reflected by extreme conditional item probabilities (as in -->
<!-- exactly 0 or 1). Such boundary parameter estimates could indicate that -->
<!-- the solution is invalid [@geiser_data_2012]. Boundary parameter -->
<!-- estimates can also happen with continuous indicators. For instance, if -->
<!-- we have a zero-inflated normal distribution and a two class solution, -->
<!-- one class might have the mean of zero and its standard deviation cannot -->
<!-- be determined since there is little variance. This too could be a sign -->
<!-- of an invalid solution, warn that too many classes were extracted, or -->
<!-- indicate a local optimum [@geiser_data_2012]. -->

@@

<!-- The total number of parameters scales with the number of estimated classes. -->
<!-- Consequently, latent class analyses have a potentially very high number of parameters. -->
<!-- As any of these parameters could be mis-specified, -->
<!-- it is important to consider alternative model specifications. -->
<!-- One way to obtain relative evidence in favor of different models is by using BIC weights, obtained by the function `tidySEM::ic_weights()` [@wagenmakersAICModelSelection2004]. -->

<!-- From sinhaPractitionerGuideLatent2021: -->
<!-- Testing for number of classes. In addition to indexing model fit, there are tests comparing a model with k classes to one with k-1 classes. Developed by Lo, Mendell and Rubin (36) based on work by Vuong (54) the VLMR test assumes multivariate normality, and it is not clear how sensitive the p-value is to violation of that assumption. It is also possible to use a bootstrapped p-value, although its validity outside of normally distributed data remains unknown.(55) In a Monte Carlo simulation of LCA using only categorical indicators, Nylund and colleagues found that the bootstrapped test (BLMR) consistently outperformed the simple Lo-Mendell-Rubin test (LMR).(34) Again in simulation studies with latent profile analysis (continuous indicators), Tein and colleagues found that the BLMR had higher statistical power than LMR, although both tests performed well at detecting the true class.(56) In our practice, across multiple analyses with real-life data consisting of a mixture of categorical and continuous indicators, we have found that the BLMR consistently favours k classes over k - 1 class to the point of being of limited value. -->
Class enumeration refers to the process of determining the appropriate number of classes.
It is approached differently for exploratory versus confirmatory LCA.
In exploratory LCA, the set of models typically consists of a sequence from 1 to the maximum number of classes $K$. 
The number $K$ may be chosen a-priori or dictated by the data.
It is sensible to limit $K$ to the maximum number of classes that result in valid and interpretable solutions (also see Classification Diagnostics).
Above a certain number of classes,
convergence problems typically arise, 
or the proportion of cases assigned to the smallest classes may be low relative to the number of class-specific parameters.
In Bayesian LCA, it has been observed that, when the estimated number of classes exceeds the true number,
the proportion of cases assigned to the excessive classes converges to zero as the sample size increases [@rousseauAsymptoticBehaviourPosterior2011].
It is not known whether this also happens in frequentist LCA;
nonetheless, near-zero class memberships should be taken into account when determining the maximum $K$.

@@

<!-- In exploratory LCA, a sequence of models is fitted to the data -->
<!-- with each additional model estimating one more class than the previous -->
<!-- model. These models are then compared and the best solution is selected -->
<!-- as the final class solution. In some cases, prior theory can inform the -->
<!-- researcher about the number of classes to expect. Even in such -->
<!-- confirmatory LCA cases, it is nonetheless useful to know if the -->
<!-- theoretical model is markedly better than those with differing numbers -->
<!-- of classes. Therefore, it may always be useful to compare different -->
<!-- class solutions. -->

@@

As there are no absolute fit indices for LCA,
information criteria are typically used to assess relative fit.
Information criteria are computed from the $-2LL$,
which is the most basic fit measure.
Information criteria add a penalty for model complexity to the $-2LL$, and thus balance fit and complexity.
The lower
the value of an information criterion, the better the overall fit of the
model.
The original information criterion is the Akaike Information Criterion (AIC),
computed as $-2LL+2p$, where $p$ is the number of parameters.
The Bayesian Information Criterion (BIC) multiplies the penalty for complexity with a function of the sample size, $p \ln(n)$,
such that complex models are penalized more in smaller samples.
Finally,
the so-called *sample size adjusted* BIC (saBIC) implements a different ad-hoc penalty for sample size,
$\ln(\frac{n + 2}{24})$.
Although a simulation study showed that the saBIC performed well for class enumeration,
its penalty is not based on theory [REF].
BIC also performs well and is theoretically substantiated;
as such, it may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
All three ICs are available in the output of `table_fit()`.

>>>>

---
title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "c.j.vanlissa@gmail.com"
  - name          : "Mauricio Garnier-Villarreal"
    affiliation   : "3"
    corresponding : no   
    email         : "m.garniervillarreal@vu.nl"
  - name          : "Daniel Anadria"
    affiliation   : "1"
    corresponding : no
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "d.anadria@uu.nl"
affiliation:
  - id            : "1"
    institution   : "Utrecht University, Methodology & Statistics"
  - id            : "2"
    institution   : "Open Science Community Utrecht"
  - id            : "3"
    institution   : "Vrije Universiteit Amsterdam, Sociology"
author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands
abstract: |
  Latent class analysis (LCA) refers to a family of techniques for identifying groups
  in data based on a parametric model.
  Examples include mixture models, LCA with ordinal
  indicators, and latent class growth analysis.
  Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of LCA,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package whose backend relies on `OpenMx`.
keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
wordcount         : "9001"
bibliography      : "mixture.bib"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
class             : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: 72
#bibliography: references.bib
---

@@

Despite the popularity of LCA, two challenges impede broader adoption
and correct application of the method. The first challenge is that most
existing software that implements these models is commercial and
closed-source. The second challenge is that there is a lack of standards
for estimating and reporting LCA, and some myths and misunderstandings
have been perpetuated in prior literature. This introduces a risk of
misapplications of the technique, and complicates manuscript review and
assessment of the quality of published studies. The present paper seeks
to address these challenges by introducing updated guidelines for
estimating and reporting LCA, based on current best practice.
Furthermore, meeting the increasing demand for open source research
software, this paper introduces new functionality in the R-package
`tidySEM` [@vanlissaTidySEMTidyStructural2022] to perform LCA using free
open source software (FOSS).

@@

There are several compelling reasons to prefer free open source software
(FOSS) instead. FOSS is freely available for anyone to use, modify, and
distribute, which means that scientists can access and use a wide range
of powerful tools without financial impediments. This is invaluable for
reducing gatekeeping of researchers with limited budgets, such as those
from developing countries. Furthermore, FOSS promotes collaboration
within the scientific community, as its users can contribute to the
development, maintenance, and support of the software. Finally, FOSS
promotes transparency and reproducibility: Because the source code for
FOSS is accessible, researchers can review and verify the methods and
algorithms used, which helps to ensure the integrity and reliability of
research findings. Overall, FOSS has the potential to significantly
improve the efficiency and effectiveness of scientific research, while
also promoting transparency, collaboration, and inclusion within the
scientific community.

Notable FOSS for LCA includes mclust [@scrucca_mclust_2016], depmixS4
[@depmixS4], and OpenMx [@neale_openmx_2016]. A crucial limitation of
these existing packages is that some have limited functionality (e.g.,
Mclust), or implement specific innovations (e.g., depmixS4). OpenMx is
unique in that, like Mplus, it implement LCA in the context of a fully
featured structural equation modeling framework. Remaining obstacles to
the use of `OpenMx` is that it is not very user-friendly and its
functionality for LCA is poorly documented.

@@

A common type of LCA is the *finite Gaussian mixture model*; a
univariate analysis where the observed distribution of a single
continuous variable is assumed to result from a mixture of a known
number of normal (Gaussian) distributions. The parameters of a finite
Gaussian mixture model are the means and variances of these underlying
normal distributions. The analysis of shoe sizes presented earlier is a
canonical example of this type of analysis. In the multivariate case,
with more than one indicator variable, the parameters of a mixture model
are the means, variances, and covariances between the indicators (which
can be standardized to obtain correlations). These parameters can be
estimated freely, or set to be constrained across classes
[@collins_latent_2009].

@@

A longitudinal extension of LCA is the Hidden Markov Model (HMM, also
called Latent Transition Analysis), which estimates the change in class
membership over time with the markov assumption over time. The markov
assumption states that the class membership at time $t$ is only affected
by the class membership at time $t-1$. The structural part of the model
is comprise of the initial states (proportion of the sample in each
class at the first time point), and transition probabilities
(probabilities to switch class membership across time). The transition
probabilities are estimated as multinomial logistic regressions between
class memberships [@vermunt_longitudinal_2010].

@@

Finally, LCA can be used to deal with data which violate certain
assumptions. For example, zero-inflated and hurdle models are special
cases of LCA for dealing data with a very high proportion of zeros
(extreme right skew) [@baughmanMixtureModelFramework2007]. These models
estimate one class to explain the "true zeros", and another class with a
different distribution to explain all other values, which can include
zero. Note that this LCA has different models for both classes, and thus
falls beyond the scope of this paper.
<!-- CJ: This has been said elsewhere: -->
<!-- In fact, LCA assumes the population -->
<!-- distribution is a non-normal mixture of $K$ normal distributions, and it -->
<!-- can recover the value of $K$, i.e. generate a $K$-class solution. -->
<!-- CJ: This example is not perfect, we need another example still.. -->
For instance, the aforementioned study on marine plastic noted that many
prior studies had treated the distribution of particles as
(multivariate) normal, despite clear deviations from normality
[@alkema_maximizing_2022]. A mixture of two normal distributions with
distinct variances described the observed distribution better.
<!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical. We encourage researchers to be agnostic about the **true** nature of the latent variable. And focus on the approach that fits the research question. -->
<!-- CJ: Why would we encourage researchers to be agnostic about the true...? -->

<!-- ### Controlling for covariates -->

<!-- CJ: I'm not sure we should call this a use case... in principle, -->

<!-- any time you're doing some kind of LCA you may want to predict group membership or predict something from group membership... -->

<!-- An extension of LCA is that containing covariates which can be used to -->

<!-- predict class membership. In this approach, we not only model the latent -->

<!-- class variable based on indicators, but we also relate the class -->

<!-- membership to other explanatory variables (@vermunt_latent_2017). An -->

<!-- example of using covariates comes from @nozadi_moderating_2016 who -->

<!-- applied LCA to identify the probability of children's membership to an -->

<!-- anxiety class. The authors tested several covariates including -->

<!-- children's age, sex, and accuracy scores. Age and sex were not found to -->

<!-- be related to the children's latent class membership, hence these -->

<!-- covariates were excluded from the analysis. Accuracy scores, however, -->

<!-- were related to probabilities of being in anxiety and attention-anxiety -->

<!-- classes. Therefore this covariate was kept as a valuable predictor. -->

<!-- Another example comes from @van_lissa_divergence_2015 who included sex  -->

<!-- as a predictor of the developmental trajectories and class membership in  -->

<!-- the previously discussed adolescent empathy development study. The rationale -->

<!-- was that adding sex as a covariate would account for known sex differences  -->

<!-- in empathy development, and lead to a model with lower bias. -->

<!-- ### Distal outcomes -->

<!-- CJ: I'm not sure we should call this a use case... in principle, -->

<!-- any time you're doing some kind of LCA you may want to predict group membership or predict something from group membership... -->

<!-- When our interest is the prediction of one or more outcomes, LCA can be used  -->

<!-- to construct latent classes as categorical predictors. @lanza_latent_2013  -->

<!-- demonstrated how LCA can be used to classify adolescents into depression  -->

<!-- classes, and subsequently these classes can be used to predict smoking, grades, -->

<!-- and delinquency. The study showed that the outcomes predicted by class  -->

<!-- membership can be binary (regular smoking), continuous (grades) or  -->

<!-- count (delinquency). In the previously discussed study of Belgian adolescent -->

<!-- identities, @maene_perceived_2022 first constructed the latent variable -->

<!-- representing different multiple identification strategies. Subsequently,  -->

<!-- the most likely class membership based on posterior class probability was -->

<!-- used as an "observed" categorical variable for prediction of depressive symptoms. -->

@@

<!-- Any variables that of class `character` or `factor` must be recoded to ordinal. -->

<!-- Nominal variables must be recoded to multiple ordinal dummy variables, e.g., using the function `mx_dummies()`.  -->

LCA can accommodate variables of different measurement levels:
continuous, binary, nominal, and ordinal. Numeric variables should have
an interval measurement level, such that it is sensible to estimate
means and (co)variances. For the remaining three measurement levels,
some preprocessing is required. The exact data types accepted by
`tidySEM` at the time of writing are `numeric` and `ordered`. To
correctly specify ordered variables such as Likert scales, use the
function `OpenMx::mxFactor()`. Binary variables, such as presence or
absence of some symptom, are also treated as ordinal, with one category
coded as zero and the other as one. Nominal variables are converted to
binary indicators of group membership via dummy coding, for example,
using `mx_dummies()`.

@@

Previous work has emphasized the importance of examining the pattern of
missing data and reporting how missingness was handled
[@van_de_schoot_grolts-checklist_2017]. While we concur that
investigating missingness is due diligence, it is important to emphasize
that missingness is adequately handled by default in `OpenMx`, which is
the computational backend of `tidySEM`. Its Full Information Maximum
Likelihood (FIML) estimator makes use of all available information
without imputing missing values, by estimating the model based on all
available information for each subject
[@endersRelativePerformanceFull2001; @enders_applied_2022]. FIML is a
best-practice solution for handling missing data; at least on par with
multiple imputation [@lee_comparison_2021]. Multiple imputation is less
suitable to LCA for two reasons. First, because LCA methods are often
computationally intensive, and conducting them on several imputed
datasets multiplies the load. Second, because many multiple imputation
methods assume normality; an assumption likely to be violated when data
come from a heterogeneous population.

@@

<!-- In the general sense, the class-specific model can be any structural equation model. -->

<!-- Its parameters may include means, (co)variances, regression coefficients, factor loadings, and thresholds for ordinal variables. -->

<!-- All of these parameters can be freely estimated, -->

<!-- constrained to be equal across classes, -->

<!-- or fixed to a known value. -->

<!-- The specific types LCA introduced in the Taxonomy all have different parameters. -->

<!-- For example, LPA typically assumes class-specific means, -->

<!-- equal variances across classes, and zero covariances (i.e., indicators are independent after accounting for latent class membership), see `mx_profiles()`. -->

<!-- By contrast, LCA with ordinal indicators assumes free thresholds across classes, -->

<!-- where each threshold reflects a proportion of responses within a particular response category, see `mx_lca()`. -->

<!-- Like LPA, LCGA and GMM typically have free means across classes, -->

<!-- but with respect to the latent growth variables instead of the observed variables, see `mx_growth_mixture()`. -->

<!-- The distinction between LCGA and GMM is whether variances of the latent growth variables are constrained to zero, or freely estimated (either across or within classes). -->

<!-- either using `lavaan` model syntax or using `OpenMx` RAM specification, -->

<!-- and supplied to the general function `mx_mixture()`. -->

<!-- The resulting model may have -->

<!-- HIER -->

<!-- For latent profile analysis, the function -->

<!-- `tidySEM::mx_profiles(classes, variances, covariances)` largely -->

<!-- automates this process. The argument `classes` indicates which class -->

<!-- solutions should be estimated (e.g., 1 through 6). The argument -->

<!-- `variances` specifies whether variances should be `"equal"` or -->

<!-- `"varying"` across classes. The argument `covariances` specifies whether -->

<!-- covariances should be constrained to `"zero"`, `"equal"` or `"varying"` -->

<!-- across classes. The means are free to vary across classes by default, -->

<!-- although the more general function `tidySEM::mx_mixture()` could be used -->

<!-- to circumvent this. After all models have been estimated, the function -->

<!-- `tidySEM::table_fit()` can be used to obtain a model fit table suitable -->

<!-- for determining the optimal model according to best practices. Note, -->

<!-- however, that this table does not include the bootstrapped likelihood -->

<!-- ratio test (BLRT) by default, because this test is very computationally -->

<!-- expensive. It is recommended to use the function `tidySEM::BLRT()` to -->

<!-- compare a shortlist of likely candidate models based on other fit -->

<!-- indices. More on fit indices can be found in the Model Fit Indices -->

<!-- subsection of this paper. -->

@@

<!-- A problem which can occur is that of inadmissible solutions. With binary -->

<!-- indicators, LCA is modeling a cross-table with all the predictors. The -->

<!-- problem with such cross-tables is that they will often contain empty -->

<!-- cells, i.e. combinations of responses that never occur together. This -->

<!-- problem is reflected by extreme conditional item probabilities (as in -->

<!-- exactly 0 or 1). Such boundary parameter estimates could indicate that -->

<!-- the solution is invalid [@geiser_data_2012]. Boundary parameter -->

<!-- estimates can also happen with continuous indicators. For instance, if -->

<!-- we have a zero-inflated normal distribution and a two class solution, -->

<!-- one class might have the mean of zero and its standard deviation cannot -->

<!-- be determined since there is little variance. This too could be a sign -->

<!-- of an invalid solution, warn that too many classes were extracted, or -->

<!-- indicate a local optimum [@geiser_data_2012]. -->

@@

<!-- The total number of parameters scales with the number of estimated classes. -->

<!-- Consequently, latent class analyses have a potentially very high number of parameters. -->

<!-- As any of these parameters could be mis-specified, -->

<!-- it is important to consider alternative model specifications. -->

<!-- One way to obtain relative evidence in favor of different models is by using BIC weights, obtained by the function `tidySEM::ic_weights()` [@wagenmakersAICModelSelection2004]. -->

<!-- From sinhaPractitionerGuideLatent2021: -->

<!-- Testing for number of classes. In addition to indexing model fit, there are tests comparing a model with k classes to one with k-1 classes. Developed by Lo, Mendell and Rubin (36) based on work by Vuong (54) the VLMR test assumes multivariate normality, and it is not clear how sensitive the p-value is to violation of that assumption. It is also possible to use a bootstrapped p-value, although its validity outside of normally distributed data remains unknown.(55) In a Monte Carlo simulation of LCA using only categorical indicators, Nylund and colleagues found that the bootstrapped test (BLMR) consistently outperformed the simple Lo-Mendell-Rubin test (LMR).(34) Again in simulation studies with latent profile analysis (continuous indicators), Tein and colleagues found that the BLMR had higher statistical power than LMR, although both tests performed well at detecting the true class.(56) In our practice, across multiple analyses with real-life data consisting of a mixture of categorical and continuous indicators, we have found that the BLMR consistently favours k classes over k - 1 class to the point of being of limited value. -->

Class enumeration refers to the process of determining the appropriate
number of classes. It is approached differently for exploratory versus
confirmatory LCA. In exploratory LCA, the set of models typically
consists of a sequence from 1 to the maximum number of classes $K$. The
number $K$ may be chosen a-priori or dictated by the data. It is
sensible to limit $K$ to the maximum number of classes that result in
valid and interpretable solutions (also see Classification Diagnostics).
Above a certain number of classes, convergence problems typically arise,
or the proportion of cases assigned to the smallest classes may be low
relative to the number of class-specific parameters. In Bayesian LCA, it
has been observed that, when the estimated number of classes exceeds the
true number, the proportion of cases assigned to the excessive classes
converges to zero as the sample size increases
[@rousseauAsymptoticBehaviourPosterior2011]. It is not known whether
this also happens in frequentist LCA; nonetheless, near-zero class
memberships should be taken into account when determining the maximum
$K$.

@@

<!-- In exploratory LCA, a sequence of models is fitted to the data -->

<!-- with each additional model estimating one more class than the previous -->

<!-- model. These models are then compared and the best solution is selected -->

<!-- as the final class solution. In some cases, prior theory can inform the -->

<!-- researcher about the number of classes to expect. Even in such -->

<!-- confirmatory LCA cases, it is nonetheless useful to know if the -->

<!-- theoretical model is markedly better than those with differing numbers -->

<!-- of classes. Therefore, it may always be useful to compare different -->

<!-- class solutions. -->

@@

As there are no absolute fit indices for LCA, information criteria are
typically used to assess relative fit. Information criteria are computed
from the $-2LL$, which is the most basic fit measure, testing the
overall model fit if the model properly represents the data. Information
criteria add a penalty for model complexity to the $-2LL$, and thus
balance fit and complexity. The lower the value of an information
criterion, the better the overall fit of the model. The original
information criterion is the Akaike Information Criterion (AIC),
computed as $-2LL+2p$, where $p$ is the number of parameters. The
Bayesian Information Criterion (BIC) multiplies the penalty for
complexity with a function of the sample size, $p \ln(n)$, such that
complex models are penalized more in smaller samples. Finally, the
so-called *sample size adjusted* BIC (saBIC) implements a different
ad-hoc penalty for sample size, $\ln(\frac{n + 2}{24})$. Although a
simulation study showed that the saBIC performed well for class
enumeration, its penalty is not based on theory [REF]. BIC also performs
well and is theoretically substantiated; as such, it may be the most
appropriate information criterion to use for model comparison
[@nylund-gibson_ten_2018; @masyn_latent_2013]. All three ICs are
available in the output of `table_fit()`.

Information criteria may occasionally contradict each other, so it is
important to identify a suitable strategy to reconcile them. One option
is to select a specific fit index, such as the BIC, before seeing the
results. This choice can be preregistered to avoid concerns about cherry
picking. Another option is to always prefer the most parsimonious model
that has best fit according to any of the available fit indices. Other
approach is to select several competing model based on the ICs, and
evaluate the theoretical intepretation. Yet another option is to
incorporate information from multiple fit indices using the analytic
hierarchy process [@akogul_comparison_2016], which is implemented in
`tidyLPA` [@rosenbergTidyLPAPackageEasily2018]. Finally, one can use a
scree plot to determine the inflection point at which additional classes
contribute little to further decreases of the ICs
[@nylund-gibson_ten_2018]. The preferred number of classes is at the
inflection point where the slope of the curve is clearly leveling off.
Adding this many classes results in substantial decreases of the IC, and
adding further classes results in relatively small decreases.
